---
title: "Main analysis"
author: "Freddie  J. Heather"
date: "01/08/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Species population trends

## Reading in raw data

```{r}
data_raw <- 
  read_csv("data/input_data.csv") %>% 
  pivot_longer(cols = `1992`:`2020`,
               names_to = "year",
               values_to = "count")

# Biogeography for each species
spp_traits <- 
  read_csv("data/spp_traits.csv") %>% 
  select(species_name_update = SPECIES_NAME,
         Biogeog = Biogeog2)

# All species 
included_spp <- 
  spp_traits %>% 
  pull(species_name_update) %>%
  unique()

# State for each site code
state_tab <- 
  read_csv("data/states_tab.csv") %>% 
  select(site_code, 
         State_new = State)

```

## Filling in gaps


```{r}

# Interpolation
data_interp <-
  data_raw %>% 
  group_by(site_code, species_name_update) %>% 
  mutate(count_interp = zoo::na.approx(count, na.rm = F)) %>% 
  mutate(latitude = round(latitude),
         longitude = round(longitude))

# Extrapolation
data_clean <- 
  data_interp %>%
  group_by(site_code, species_name_update) %>%
  mutate(count_extrap = count_interp) %>%
  fill(count_extrap, .direction = "downup") %>%
  ungroup()

```

## Mean counts per lat-lon grid cell

```{r}

# Mean within grid cells
data_latlon_means <- 
  data_clean %>% 
  group_by(latitude, longitude, MPA, Taxon, species_name_update, year, State) %>% 
  summarise(mean_count_latlon = mean(count_extrap, na.rm = T), 
            .groups = "drop")

# species mean of within that grid cell
data_latlon_means_perspp <- 
  data_latlon_means %>% 
  group_by(Taxon, species_name_update, year) %>% 
  summarise(mean_count_latlon_perspp = mean(mean_count_latlon, na.rm = T), 
            .groups = "drop")

```

## Logging mean count values

```{r}

# min value to be used when mean count is zero 
min_vals <- 
  data_latlon_means_perspp %>% 
  filter(year >= 2008) %>% 
  group_by(species_name_update) %>% 
  summarise(min_latlon_perspp = min(mean_count_latlon_perspp[mean_count_latlon_perspp != 0], na.rm = T),
            .groups = "drop") 

# log transform, use (min count for the species)/2 when no mean count is available
data_latlon_means_perspp_log <- 
  data_latlon_means_perspp %>% 
  filter(year >= 2008) %>%
  left_join(min_vals, by = "species_name_update") %>%
  mutate(log_mean_count_latlon_perspp = case_when(mean_count_latlon_perspp == 0 ~ log(min_latlon_perspp/2), 
                                    TRUE            ~ log(mean_count_latlon_perspp))) 

```

## Slope values of population trends

```{r}

# get the slope of the last 2008-2020 data
latlon_means_perspp_slopes <-
  data_latlon_means_perspp_log %>% 
  filter(log_mean_count_latlon_perspp != Inf) %>% 
  group_by(species_name_update) %>% 
  nest() %>% 
  mutate(lm =  map(data, ~lm(.x$log_mean_count_latlon_perspp ~ as.numeric(.x$year))),
         slope = map_dbl(lm, ~.x$coefficients[2]), 
         r2 = map_dbl(lm, ~summary(.x)$r.squared),
         adj_r2 = map_dbl(lm, ~summary(.x)$adj.r.squared)) %>% 
  select(-c(data, lm, r2, adj_r2))


# 2008-10 vs 2018-2020 mean counts
latlon_means_perspp_changeratio <-
  data_latlon_means_perspp_log %>% 
  mutate(decade = case_when(year %in% 2008:2010 ~ "first",
                            year %in% 2018:2020 ~ "second",
                            TRUE ~ NA_character_)) %>% 
  group_by(species_name_update, decade) %>%
  summarise(decade_mean = mean(mean_count_latlon_perspp, na.rm = T)) %>% 
  filter(!is.na(decade)) %>% 
  pivot_wider(names_from = decade,
              values_from = decade_mean) %>% 
  mutate(change_ratio = second/first) %>% 
  select(-c(second, first))

# Number of counts per species
count_spp <-
  data_clean %>% 
  filter(count > 0) %>% 
  group_by(species_name_update) %>% 
  summarise(total_count = n())

output_poptrend <-
  full_join(latlon_means_perspp_slopes, latlon_means_perspp_changeratio) %>% 
  left_join(count_spp) %>% 
  arrange(species_name_update) %>% 
  mutate_if(is.numeric, list(~na_if(., Inf))) %>% 
  mutate_if(is.numeric, list(~na_if(., -Inf)))

write_csv(output_poptrend, "output/spp_poptrend.csv")

```

# Analysis 2

## Logging raw count values

```{r}

# min values of the original counts to be used to log zero
min_vals_raw <- 
  data_clean %>% 
  group_by(species_name_update, site_code) %>% 
  summarise(min = min(count[count != 0], na.rm = T),
            .groups = "drop") 

# log of count within cells
data_logcount <- 
  data_clean %>% 
  select(species_name_update, year, count, site_code, latitude, longitude, MPA) %>% 
  left_join(min_vals_raw) %>%
  mutate(log_count = case_when(count == 0 ~ log(min/2), 
                               TRUE       ~ log(count))) 
```

## Standardisation

Log counts are standardised to the mean value for the given site and species.

```{r}

# mean logcount - to be used as difference from raw count within spp within cells
data_logcount_mean <-
  data_logcount %>% 
  group_by(species_name_update, site_code, latitude, longitude) %>% 
  summarise(mean_logcount = mean(log_count, na.rm = T))

# logcount minus mean(log count) within spp within cells
data_logcounts_std <- 
  data_logcount %>% 
  left_join(data_logcount_mean) %>% 
  mutate(std_logcount = log_count - mean_logcount) %>% 
  group_by(latitude, longitude, MPA, species_name_update, year) %>% 
  summarise(mean_std_logcount = mean(std_logcount, na.rm = T), 
            .groups = "drop") %>% 
  mutate(mean_std_logcount = ifelse(is.nan(mean_std_logcount), NA, mean_std_logcount))

# mean of standardised log counts per species per year
data_logcounts_std_sppyear <- 
  data_logcounts_std %>% 
  group_by(species_name_update, year) %>% 
  summarise(mean_std_logcount_sppyear = mean(mean_std_logcount, na.rm = T), 
            .groups = "drop") %>%   
  mutate(mean_std_logcount_sppyear = ifelse(is.nan(mean_std_logcount_sppyear), NA, mean_std_logcount_sppyear)) %>% 
  filter(year >= 2008) # only looking at post-2007 data

```


```{r}

# Spearmans rank of the standardised log counts
output_spearmans <- 
  data_logcounts_std_sppyear %>% 
  group_by(species_name_update) %>%
  na.omit() %>% 
  nest() %>% 
  mutate(n_row = map_dbl(data, ~nrow(.x))) %>% 
  filter(n_row > 5) %>% 
  mutate(rho = map_dbl(data, ~cor(as.numeric(.x$year), .x$mean_std_logcount_sppyear, method = "spearman")),
         pval = map_dbl(data, ~cor.test(as.numeric(.x$year), .x$mean_std_logcount_sppyear, method = "spearman")$p.val)#,
         #pval = map_dbl(data, ~cor.test(as.numeric(.x$year), rank(.x$mean_std_logcount2))$p.val)
  ) %>% 
  mutate(sig = case_when(pval <= 0.001 ~ "***",
                         pval <= 0.01  ~ "**",
                         pval <= 0.05  ~ "*",
                         TRUE ~ "")) %>% 
  mutate(direction = case_when(sig != "" & rho > 0 ~ "up",
                               sig != "" & rho < 0 ~ "down",
                               TRUE ~ "")) %>% 
  select(-data)

write_csv(output_spearmans, "output/output_spearmans.csv")

```

# Plotting

## Data for plotting


```{r}

# minimum values including interpolated/extrapolated values
# Note: this could be less than zero if interpolated between 0 and 1.
min_vals_interp <- 
  data_clean %>% 
  group_by(species_name_update, site_code) %>% 
  summarise(min = min(count_extrap[count_extrap != 0], na.rm = T),
            .groups = "drop")

# replace zero values by min/2 before logging
data_log_interp <- 
  data_clean %>% 
  left_join(min_vals_interp) %>% 
  mutate(log_extrap = case_when(count_extrap == 0 ~ log(min/2),
                                TRUE ~ log(count_extrap)))

# log value for 2008 for standardisation
val_2008 <- 
  data_log_interp %>% 
  select(site_code, species_name_update, year, log_extrap) %>% 
  filter(year == 2008) %>% 
  distinct() %>% 
  rename(log_extrap_2008 = log_extrap) %>% 
  select(-year)

# Log count - 2008 log count (difference of logs)
data_log_interp_std <- 
  data_log_interp %>% 
  left_join(val_2008) %>% 
  mutate(log_extrap_std = log_extrap - log_extrap_2008)

```
## Mean standardised log of extrapolated counts

```{r}

# adding biogeog and State information
data_log_interp_std_step1 <- 
  data_log_interp_std %>% 
  left_join(spp_traits) %>% 
  filter(species_name_update %in% included_spp) %>% 
  select(-State) %>% 
  left_join(state_tab) 

# post 2007 means
data_log_interp_std_means <- 
  data_log_interp_std_step1 %>% 
  filter(year >= 2008) %>% 
  group_by(Taxon, State_new, Biogeog, species_name_update, year) %>% 
  summarise(mean_log_extrap_std = mean(log_extrap_std, na.rm = T), .groups = "drop")


```

## P1 = Biogeog trends per State

```{r}

p1_data <-
  data_log_interp_std_means %>% 
  group_by(State_new, Biogeog, year) %>% 
  summarise(mean_log_extrap = 
              mean(mean_log_extrap_std, na.rm = T), .groups = "drop")

write_csv(p1_data, "output/p1_data.csv")
```

## P2 = Taxon trends by State

```{r}

p2_data <-
  data_log_interp_std_means %>% 
  group_by(State_new, Taxon, year) %>% 
  summarise(mean_log_extrap = mean(mean_log_extrap_std, na.rm = T), 
            .groups = "drop")

write_csv(p2_data, "output/p2_data.csv")
```

## P3 = Biogeog trends by 5 degs lat

```{r}

lats <- 
  data_clean %>% 
  select(site_code, latitude, longitude) %>% 
  distinct()

lat_bins <- 
  tibble(lat_bins = seq(-45, 0, by = 5)) %>% 
  mutate(bin_num = 1:length(lat_bins))

lat_5degs <- 
  data_clean %>% 
  select(site_code, latitude, longitude) %>% 
  distinct() %>% 
  mutate(lat_bin = lat_bins$lat_bins[.bincode(latitude + 5, lat_bins$lat_bins)])

p3_data <-
  data_log_interp_std_step1 %>% 
  left_join(lat_5degs) %>% 
  group_by(lat_bin, Biogeog, Taxon, species_name_update, year ) %>% 
  filter(year  >= 2008) %>% 
  summarise(mean_val = mean(log_extrap_std, na.rm = T)) %>% 
  group_by(lat_bin, Biogeog, year) %>% 
  summarise(mean = mean(mean_val, na.rm = T))

write_csv(p3_data, "output/p3_data.csv")

```

